A family of accessibility measures derived from spatial interaction principles

### 2025-08-28 paper_plos/paper_plos.qmd
rafa up to page 10

@@

In the early nineteenth century, rapid urbanisation driven by industrialisation expanded cities across Great Britain, Northern Europe, and North America. This expansion led to traffic congestion and the emergence of a transportation planning paradigm focused primarily on mobility. Within this new practice, access to destinations was treated as a by-product of movement. After World War II, major investments in automobile and transportation infrastructure cemented this mobility-oriented model, fostering flight from urban congested cities, lower-density sprawl, car-dependent development and entrenching automobility in planning practice [@miller_collaborative_2011; @lavery_driving_2013]. Despite continued road and highway expansion, this automobility monoculture has proven ineffective at reducing travel costs or environmental burdens, and has not clearly improved people’s ability to reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].

@@

More specifically, the wide range of accessibility definitions--with novel methods being more sophisticated but less intuitive [@kapatsila_resolving_2023]--can further hinder practical uptake [@vanweeAccessible2016]. Geurs and van Wee [@geursAccessibilityEvaluationLanduse2004] classify accessibility measures into four categories: infrastructure-, place-, person-, and utility-based. Among place-based measures (this work's focus), variants include gravity-based [@hansen1959], cumulative opportunity [@pirie_measuring_1979], the 2 Step Floating Catchment Area (FCA) approach [@luo2003], and a variety of modifications to these approaches e.g., Enhanced 2-Step FCA [@luoEnhanced2009], 3-Stage FCA [@wan2012three], Modified 2-Step FCA [@delamater2013spatial], inverse 2-Step FCA [@wang_2sfca_2021], and n-steps FCA [@liang_novel_2024]. While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised in [@vanweeAccessible2016]: How should practitioners interpret difference in scores between modes, and how should results be communicated?

@@

Where $G$ is the gravitational constant--an empirically calibrated value that ensures the model reflects observed forces. Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's experiments of 1778 [@hutton_xxxiii_1778, p. 783] and Cavendish's 1798 experiment [@cavendish_xxi_1798] to estimate of the proportionality constant to within 1% accuracy. In sum: it took over a century from the publication of *Principia* to refine the estimate of the proportionality constant.

@@

Following Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858], research into human spatial interaction continued in different contexts. Namely, since Carey to the 1940s, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance--conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.

@@

A common feature of these early investigations of human spatial interaction is that a proportionality constant, similar to $G$ in @eq-phys-grav was never considered. Of the researchers cited above, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was presented in the following form:

@@

These early formulations (@eq-reilly, @eq-zipf) clearly reflect the influence of Newtonian gravity on human spatial interaction theory, revealing a shared mathematical structure across migration, trade, and communication models. However, none included a proportionality constant--highlighting a consistent omission of the empirical calibration necessary to convert these proportional relationships into measurable and comparable quantities.

After Carey and folowing Reilly and Zipf, it is only in Stewart [@stewartDemographicGravitationEvidence1948] that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant. <!-- If we were to ponder as of to _why_: it may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists. --> Besides awareness of preceding research (i.e., Stewart cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, enabling his formulation to be interpreted as a measurable 'demographic' force:

@@

Although Stewart’s concept of "social physics" eventually fell out of favour--due to inconsistent mathematical notation (e.g., $G$ is used as both a proportionality constant p. 34 and then later as _demographic energy_ on p. 53.) and its racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight $\mu$ of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p. 35\])--his introduction of a proportionality constant $G$ in modeling demographic force marks an important conceptual step: recognizing that moving from proportionality to equality requires empirical calibration. In other words, the addition of $G$ shifts results from being abstract indicators of potential (i.e., $\frac{\text{people}^2}{\text{distance}^{2}}$) to having interpretable units grounded in consistent, albeit still abstract, quantities (i.e., units of demographic force).

@@

$S_{i}$ in @eq-accessibility-general is a measure of the accessibility from site $i$. This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $\beta$ (a parameter that modulates the friction of cost). Today, Hansen is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to as the gravity-based accessibility measure.

@@

Therefore in practice, Hansen-type measures are ones of proportionality and are better understood as *ordinal indicators*; they rank accessibility but lack cardinal meaning or consistent units [@millerAccessibilityMeasurementApplication2018]. The continued absence of a sort of proportionality constant $G$ leaves a conceptual and practical gap in accessibility analysis: a missing link between theoretical form and empirical measurement.

# Wilson's family of spatial interaction models
While accessibility research evolved in North America with Hansen [@hansen1959], a parallel development was taking place across the Atlantic with Alan G. Wilson. Wilson's ground breaking paper [@wilson1971], advanced a general framework for spatial interaction modeling--focused on modeling flows of interaction between places. This work was not focused on the 'potential' concept as accessibility has been. Wilson [@wilson1971] formalized the general spatial interaction model through the following equation:

@@

Depending on which of the three system constraints are applied a family of spatial interaction models can be derived from @eq-phys-gravity-model. The proportionality constant $k$ is replaced with different _balancing factors_. This change in name is more useful as these factors not only preserve proportionality but also ensure that the predicted flows $T_{ij}$ align with the known constraints considered in the system. In other words, the different balancing factors adjust the model so that flows become statistical averages consistent with observed origin and/or destination data. In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$), and (3) a doubly constrained model that satisfies both.

@@

Rather than relying on a universal constant like $G$, Wilson’s models calibrate interaction flows through known empirical constraints (e.g., total system flows, or origin-side and destination-side flow totals). This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows. In short, when aspects of the systems are inputs into the balancing factors, $ij$ flows can be robustly predicted in tangible units of 'flow', units that align with the system being modelled. In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled. While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today) yield outputs that _indicate_ a proportional relationship and not one of equality nor yielding results without interpretable units.

>>>>

\textcolor{red}{Rafa: Do we have to shorthen the abstract? I actually preferred the abstract we had in the preprint. I like that the abstract mentions the toy example though, which was not done in the 1st version.}

@@

In the early twentieth century, the emergence of a transportation planning paradigm focused primarily on mobility cemented major investments in automobile and transportation infrastructure, fostering lower-density sprawl, car-dependent development and entrenching automobility in planning practice [@miller_collaborative_2011; @lavery_driving_2013]. Within this new practice, access to destinations was treated as a by-product of movement. Despite continued road and highway expansion, this automobility monoculture has proven ineffective at reducing travel costs or environmental burdens, and has not clearly improved people’s ability to reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].

@@

More specifically, the wide range of accessibility definitions--with novel methods being more sophisticated but less intuitive [@kapatsila_resolving_2023]--can further hinder practical uptake [@vanweeAccessible2016]. Geurs and van Wee [@geursAccessibilityEvaluationLanduse2004] classify accessibility measures into four categories: infrastructure-, place-, person-, and utility-based. Among place-based measures (this work's focus), variants include gravity-based [@hansen1959], cumulative opportunity [@pirie_measuring_1979], the 2 Step Floating Catchment Area (FCA) approach [@luo2003], and a variety of modifications to these approaches e.g., Enhanced 2-Step FCA [@luoEnhanced2009], 3-Stage FCA [@wan2012three], Modified 2-Step FCA [@delamater2013spatial], inverse 2-Step FCA [@wang_2sfca_2021], and n-steps FCA [@liang_novel_2024]. While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised in [@vanweeAccessible2016]: How should practitioners interpret differences in accessibility scores between modes, and how should results be communicated?

@@

Where $G$ is the gravitational constant--an empirically calibrated value that ensures the model reflects observed forces. Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's and
Cavendish's experiments in the late 1700s [@hutton_xxxiii_1778; @cavendish_xxi_1798] to estimate of the proportionality constant to within 1% accuracy. In sum: it took over a century from the publication of *Principia* to refine the estimate of the proportionality constant.

@@

Following Carey's work [@careyPrinciplesSocialScience1858], research into human spatial interaction continued in different contexts. Namely, since Carey to the 1940s, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance--conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.

@@

Of the researchers cited above, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was presented in the following form:

@@

These early formulations (@eq-reilly, @eq-zipf) clearly reflect the influence of Newtonian gravity on human spatial interaction theory, revealing a shared mathematical structure across migration, trade, and communication models.

However, a common feature of these early investigations is that none of them included a proportionality constant (similar to $G$ in @eq-phys-grav), a consistent omission of the empirical calibration necessary to convert these proportional relationships into measurable and comparable quantities. It is only in Stewart [@stewartDemographicGravitationEvidence1948] that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant. <!-- If we were to ponder as of to _why_: it may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists. --> While acknowledging predecessors like Reilly and Zipf, the physicist Stewart was likely the first author to formalize human spatial interaction using an explicit proportionality constant $G$, enabling his formulation to be interpreted as a measurable 'demographic' force:

@@

Although Stewart’s concept of "social physics" eventually fell out of favour--due to inconsistent mathematical notation (e.g., $G$ is used as both a proportionality constant p. 34 and then later as _demographic energy_ on p. 53.) and its racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p. 35\])--his introduction of a proportionality constant $G$ in modeling demographic force marks an important conceptual step: recognizing that moving from proportionality to equality requires empirical calibration. In other words, the addition of $G$ shifts results from being abstract indicators of potential (i.e., $\frac{\text{people}^2}{\text{distance}^{2}}$) to having interpretable units grounded in consistent, albeit still abstract, quantities (i.e., units of demographic force).

@@

\textcolor{red}{Rafa: there is no $\beta$ in this equation. Check text in paragraph below.}

$S_{i}$ in @eq-accessibility-general is a measure of the accessibility from site $i$. This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $f(d_{ij})$ (a function that modulates the friction of cost). Today, Hansen is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to as the gravity-based accessibility measure.

@@

Therefore in practice, Hansen-type measures are ones of proportionality and are better understood as *ordinal indicators*; they rank accessibility but lack cardinal meaning or consistent units [@millerAccessibilityMeasurementApplication2018]. The continued absence of a sort of proportionality constant $G$ leaves a conceptual and practical gap in accessibility analysis: a missing link between theoretical form and empirical measurement that hinders the interpretability of accessibility measures.

# Wilson's family of spatial interaction models
While accessibility research evolved in North America with Hansen [@hansen1959], a parallel development was taking place across the Atlantic with Alan G. Wilson. Wilson's ground breaking paper [@wilson1971], advanced a general framework for spatial interaction modeling focused on flows of interaction between places. This work was not focused on the 'potential' concept as accessibility has been. Wilson [@wilson1971] formalized the general spatial interaction model through the following equation:

@@

Depending on which of the three system constraints are applied a family of spatial interaction models can be derived from @eq-phys-gravity-model. The proportionality constant $k$ is replaced with different _balancing factors_. This change in name is more useful as these factors not only preserve proportionality but also ensure that the predicted flows $T_{ij}$ align with the known constraints considered in the system. In other words, the different balancing factors adjust the model so that flows become statistical averages consistent with observed origin and/or destination data.

In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$), and (3) a doubly constrained model that satisfies both.

@@

Rather than relying on a universal constant like $G$, Wilson’s models calibrate interaction flows through known empirical constraints (e.g., total system flows, or origin-side and destination-side flow totals). This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows. In short, when aspects of the systems are inputs into the balancing factors, $ij$ flows can be robustly predicted in tangible units of 'flow', units that align with the system being modelled. In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled. While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today) yield outputs that reflect proportional—not equal—relationships and typically lack interpretable units.
